---
title: "Training and validation rows of the neural network chain ladder"
subtitle: "The split of the learning cells by development period"
author: "Prepared for Francis"
date: "4 October 2026"
output:
  word_document: default
---

```{r setup, include = FALSE}
## after rendering to .docx, run
## python "Claude outputs/nncl-word-tables/format_word_tables.py" <file.docx>
##   720,920,880,1050,950,1050,1050,950,1050,1018 18
## which sets the borders, fonts (9 pt) and column widths of the table
knitr::opts_chunk$set(echo = FALSE)
suppressPackageStartupMessages(library(data.table))

cells <- readRDS(file.path("..", "data", "interim",
                           "nncl_synthetic_cells.rds"))
n_ay <- 20
cum <- as.matrix(cells[, paste0("cum_", 0:(n_ay - 1)), with = FALSE])

## thousands separated by a non-breaking space
num <- function(v, digits) {
  format(round(v, digits), nsmall = digits, big.mark = " ", trim = TRUE)
}
ay_range <- function(ay) {
  if (min(ay) == max(ay)) as.character(min(ay)) else
    paste(min(ay), max(ay), sep = "-")
}

## learning cells of development period j, Keras's split (the last 10% of
## the rows are the validation rows) and the homogeneous CL factors
split <- rbindlist(lapply(1:(n_ay - 1), function(j) {
  r <- which(cells$i <= n_ay - j & cum[, j] > 0)
  train <- r[seq_len(floor(0.9 * length(r)))]
  vali <- setdiff(r, train)
  factor <- function(rows) sum(cum[rows, j + 1]) / sum(cum[rows, j])
  data.table(j = j,
             learn = num(length(r), 0),
             train = num(length(train), 0),
             vali = num(length(vali), 0),
             ay_learn = ay_range(cells$i[r]),
             ay_vali = ay_range(cells$i[vali]),
             share = paste0(num(100 * sum(cum[vali, j]) / sum(cum[r, j]), 1),
                            "%"),
             f_train = num(factor(train), 5),
             f_vali = num(factor(vali), 5),
             f_learn = num(factor(r), 5))
}))
```

Neural network chain ladder on the SynthETIC portfolio. Network *j* estimates the factor from development year *j* - 1 to *j*. Its learning cells are ordered by accident year and then by the covariates; the first 90% are the training rows and the last 10% the validation rows. The volume share is the share of the cumulative payments at development year *j* - 1 that lies in the validation rows. A factor is the sum of the cumulative payments at development year *j* over the sum at development year *j* - 1, taken over the rows named.

```{r}
knitr::kable(
  split,
  format = "pipe",
  align = "rrrrccrrrr",
  col.names = c("Period j",
                "Learning cells",
                "Training rows",
                "Validation rows",
                "Accident years, learning cells",
                "Accident years, validation rows",
                "Volume share of validation rows",
                "Factor on training rows",
                "Factor on validation rows",
                "Factor on all learning cells")
)
```
