---
title: "The optimisers under the chain-ladder start"
subtitle: "Validation loss, reserve stability and training length by optimiser"
author: "Prepared for Francis"
date: "4 October 2026"
output:
  word_document: default
---

```{r setup, include = FALSE}
## after rendering to .docx, run
## python "Claude outputs/nncl-word-tables/format_word_tables.py" <file.docx>
##   1060,1020,1020,1020,1020,1400,1020,1020,1058 18
## which sets the borders, fonts (9 pt) and column widths of the tables
knitr::opts_chunk$set(echo = FALSE)
suppressPackageStartupMessages(library(data.table))

## all runs of the grid; here the training cl_start only
grid <- fread(file.path("..", "output", "tables", "04_NN-chain-ladder",
                        "synthetic", "nncl_synthetic_layers_optimisers.csv"))
grid <- grid[training == "cl_start"]
n_seeds <- uniqueN(grid$seed)

## validation loss of the chain-ladder start: the homogeneous factor of the
## training rows on the validation rows, summed over the development periods
cells <- readRDS(file.path("..", "data", "interim",
                           "nncl_synthetic_cells.rds"))
n_ay <- 20
units <- 1e6
cum <- as.matrix(cells[, paste0("cum_", 0:(n_ay - 1)), with = FALSE])
start_vali <- sum(sapply(1:(n_ay - 1), function(j) {
  r <- which(cells$i <= n_ay - j & cum[, j] > 0)
  train <- seq_len(floor(0.9 * length(r)))
  f_start <- sum(cum[r[train], j + 1]) / sum(cum[r[train], j])
  sum(((cum[r, j + 1] - f_start * cum[r, j])^2 / cum[r, j])[-train])
})) / units

## true outstanding payments: C_{i,J}(x) - C_{i,I-i}(x) of the accident
## years 2..I (the lower triangle)
true_reserve <- sum((cum[, n_ay] - cells$c_diag)[cells$i > 1]) / units

## the five optimisers of the thesis tables, in their order
optimisers <- c(sgd = "SGD", sgd_momentum = "Momentum SGD",
                rmsprop = "RMSprop", adam = "Adam", nadam = "Nadam")
## rounds halves up (the csv values are already rounded to one decimal);
## thousands separated by a non-breaking space
num <- function(v, digits) {
  format(round(v + 1e-9, digits), nsmall = digits, big.mark = " ",
         trim = TRUE)
}
optimiser_table <- function(h) {
  s <- grid[hidden == h,
            .(vali_mean = mean(loss_vali),
              vali_sd = sd(loss_vali),
              vali_min = min(loss_vali),
              vali_max = max(loss_vali),
              reserve_mean = mean(reserve),
              reserve_sd = sd(reserve),
              steps = mean(epochs_used)),
            by = optimizer]
  s <- s[match(names(optimisers), optimizer)]
  knitr::kable(
    data.frame(optimisers[s$optimizer],
               num(s$vali_mean, 1),
               num(s$vali_sd, 1),
               num(s$vali_min, 1),
               num(s$vali_max, 1),
               paste0(num(100 * (1 - s$vali_mean / start_vali), 2), "%"),
               num(s$reserve_sd, 1),
               # bias of the reserve averaged over the seeds
               paste0(num(100 * (s$reserve_mean / true_reserve - 1), 2), "%"),
               num(s$steps, 0)),
    format = "pipe",
    row.names = FALSE,
    align = "lrrrrrrrr",
    col.names = c("Optimiser",
                  "Mean validation loss",
                  "Sample standard deviation of validation loss",
                  "Minimum validation loss",
                  "Maximum validation loss",
                  "Proportion below the homogeneous chain ladder",
                  "Sample standard deviation of reserves (millions)",
                  "Bias of mean reserve (% of true reserve)",
                  "Steps used per run")
  )
}
```

Neural network chain ladder on the SynthETIC portfolio with the chain-ladder start (training `cl_start`), `r n_seeds` seeds per optimiser. Amounts are in millions. The validation loss is summed over the 19 development periods; the homogeneous chain ladder (the start of every network) scores `r num(start_vali, 1)`. The bias is the reserve averaged over the `r n_seeds` seeds less the true outstanding payments of `r num(true_reserve, 1)`, in percent of the latter; Mack's chain ladder on the same triangle has a bias of -1.53%.

## One hidden layer of 20 neurons

```{r}
optimiser_table("20")
```

## Three hidden layers of 20, 20 and 15 neurons

```{r}
optimiser_table("20-20-15")
```

## Three hidden layers of 20, 15 and 10 neurons

```{r}
optimiser_table("20-15-10")
```
