---
title: "The SynthETIC claims portfolio: data analysis and the impact of the claim features"
author: "Prepared for Francis"
date: "2 October 2026"
output:
  word_document:
    reference_docx: guide_reference.docx
    toc: true
    toc_depth: 2
    number_sections: true
    pandoc_args: ["--lua-filter=guide_tables.lua"]
toc-title: "Contents"
---

```{r setup, include = FALSE}
##########################################
#########  The SynthETIC claims portfolio: report
#########  reads the tables and figures of the two scripts in
#########  analysis/00_claim-simulation: "SynthETIC claims description.R"
#########  and "SynthETIC feature impact.R" (run them first)
##########################################

library(knitr)

opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
options(knitr.kable.NA = "")

## tables and figures of the two scripts
tab_dir <- here::here("output", "tables", "00_claim-simulation")
fig_dir <- here::here("output", "figures", "00_claim-simulation")
data_tab <- function(name) {
  read.csv(file.path(tab_dir, "data-description",
                     paste0("synthetic_data_", name, ".csv")),
           check.names = FALSE)
}
feat_tab <- function(name) {
  read.csv(file.path(tab_dir, "feature-impact",
                     paste0("synthetic_features_", name, ".csv")),
           check.names = FALSE)
}
data_fig <- function(name) {
  file.path(fig_dir, "data-description",
            paste0("SynthETIC data Fig ", name, ".png"))
}
feat_fig <- function(name) {
  file.path(fig_dir, "feature-impact",
            paste0("SynthETIC features Fig ", name, ".png"))
}

## captions in the order of the document: the position is the number
tab_caps <- c(
  simulation = "Settings of the simulation",
  portfolio = "The portfolio at the valuation date",
  accident_years = paste("Accident years: claims, claim cost and nominal",
                         "payments (true; amounts in millions, sizes in",
                         "units)"),
  claim_sizes = "Distribution of the claim size (true)",
  concentration = "Concentration of the claim cost",
  largest = "The ten largest claims (amounts in millions)",
  delays = "Distribution of the delays in years (true)",
  reported = paste("Reported claims by accident year and reporting delay T",
                   "in years (observed) and the claims not yet reported"),
  ibnyr = paste("Chain-ladder estimate of the claims not yet reported",
                "against the true number"),
  status = paste("Claim status at the valuation date and the true",
                 "outstanding (millions)"),
  censoring = "True means of the closed, open and unreported claims",
  by_delay = "Claims by reporting delay T (years)",
  by_payments = "Claims by number of payments",
  by_size = "Claims by claim size band",
  incremental = "Observed incremental payments (millions)",
  cumulative = paste("Observed cumulative payments and the true ultimate at",
                     "development year 20 (millions)"),
  development = "Development factors and payment pattern",
  calendar = "Payments by calendar year (millions)",
  late = "Payments after development year 20 (millions)",
  cl_reserves = paste("Chain-ladder reserves and Mack's sqrt(msep) against",
                      "the true outstanding (millions)"),
  f_design = "Design of the five features by level",
  f_interactions = "Designed relativities between two features that are not 1",
  f_mix = "Observed and designed share of the claims by level",
  f_dependent = "The two pairs of dependent features",
  f_cramer = "Association of the features: Cramer's V",
  f_combos = "The 480 combinations of feature levels",
  f_top = "The ten feature combinations with the largest claim cost",
  f_stability = "Mix of the levels over the 20 accident years",
  f_size = "Claim size by level (one-way)",
  f_sev = "Severity relativities against the reference level",
  f_injury_age = "Claim size by Injury Severity and Age of Claimant",
  f_tail = "The levels in the largest 1% of the claims",
  f_delays = "Notification and settlement delay by level in years (one-way)",
  f_delay_mult = "Delay multipliers against the reference level",
  f_truncation = paste("Settlement multipliers estimated on all claims and",
                       "on the claims closed at the valuation date"),
  f_pay_band = "Number of payments by claim size band",
  f_pay_level = "Number of payments by level",
  f_pay_r2 = "Number of payments: variance explained",
  f_variance = "Variance explained by the features (%)",
  f_pattern = paste("Payment pattern by level: cumulative % of the nominal",
                    "ultimate paid by development year (true)"),
  f_timing = "Payment timing by level (true)",
  f_exposure = paste("Share of the claims, of the cost and of the true",
                     "outstanding by level"),
  f_open = paste("Open and unreported claims by level and their true",
                 "outstanding (millions)"),
  f_os_injury = paste("True outstanding by accident year and Injury",
                      "Severity (millions)"),
  f_cl_factors = "Chain-ladder factors of each level's own observed triangle",
  f_cl_reserves = paste("Chain-ladder reserve of each level's own triangle",
                        "against its true outstanding (millions)"),
  f_cl_split = paste("Sum of the level reserves against the chain ladder of",
                     "the whole triangle (millions)")
)
fig_caps <- c(
  ay = "Claims, claim cost and claim size by accident year (true)",
  size = "Claim size before and after the features (density of the log)",
  lorenz = "Lorenz curves of the claim cost",
  delays = "Distribution functions of the notification and settlement delay",
  open = paste("Claims not closed at the valuation date and their true",
               "outstanding, accident years 11 to 20"),
  bars_delay = "Claims by reporting delay T (dashed: portfolio average)",
  bars_payments = paste("Claims by number of payments (15 = 15 or more;",
                        "dashed: portfolio average)"),
  triangle = paste("Incremental payments in % of the true ultimate of the",
                   "accident year; the staircase is the valuation date"),
  dev = paste("Cumulative payments by accident year in % of the true",
              "ultimate"),
  link = paste("Observed individual development factors by accident year",
               "(dashed: chain-ladder factor)"),
  calendar = paste("Payments by calendar year in millions (vertical line:",
                   "valuation date)"),
  reserves = paste("True outstanding and chain-ladder reserves with two",
                   "sqrt(msep) by accident year"),
  f_mix = "Share of the claims by level",
  f_pairs = paste("Number of claims in the two pairs of dependent features",
                  "(white: not possible by design)"),
  f_mix_ay = paste("Mean and range over the accident years of the level's",
                   "share of the claims and of the nominal ultimate"),
  f_size = paste("Claim size by level: median with quartiles and mean",
                 "(vertical line: portfolio median)"),
  f_sev = "Severity relativities: designed, one-way and all else equal",
  f_injury_age = paste("Median claim size in thousands (number of claims) by",
                       "Injury Severity and Age of Claimant"),
  f_delay_mult = "Delay multipliers: designed, one-way and all else equal",
  f_np = paste("Mean number of payments against the claim size by Injury",
               "Severity (grey: designed mean)"),
  f_variance = paste("Variance explained by each feature: drop-one (bars)",
                     "and one-way (circles)"),
  f_pattern = paste("Payment pattern of each level (black) against the",
                    "portfolio (grey)"),
  f_shares = paste("Share of the claims, of the nominal ultimate and of the",
                   "true outstanding by level"),
  f_factors = paste("Chain-ladder factors of each level's own triangle",
                    "(black) against the portfolio (grey)"),
  f_cl_error = paste("Error of each level's chain-ladder reserve in % of its",
                     "true outstanding, with one sqrt(msep) to each side")
)
tab_ref <- function(label) paste("Table", match(label, names(tab_caps)))
fig_ref <- function(label) paste("Figure", match(label, names(fig_caps)))
tab_cap <- function(label) paste0(tab_ref(label), ": ", tab_caps[[label]])
fig_cap <- function(label) paste0(fig_ref(label), ": ", fig_caps[[label]])

## numbers as text: thousands separator, fixed decimals
num <- function(x, digits = 0) {
  ifelse(is.na(x), NA,
         formatC(x, format = "f", digits = digits, big.mark = ","))
}
## each number with the decimals its size needs (tables of mixed quantities)
smart <- function(x) {
  ifelse(is.na(x), NA,
         ifelse(x == round(x), num(x, 0),
                ifelse(abs(x) >= 100, num(x, 1),
                       ifelse(abs(x) >= 1, num(x, 2),
                              formatC(x, format = "fg", digits = 3)))))
}
## a table: column names, decimals per column, caption
show <- function(d, label, names, digits = 1, align = NULL) {
  if (is.null(align)) align <- ifelse(sapply(d, is.numeric), "r", "l")
  kable(d,
        col.names = names,
        digits = digits,
        align = align,
        format.args = list(big.mark = ","),
        caption = tab_cap(label),
        row.names = FALSE)
}

#########  load the tables

portfolio <- data_tab("t01_portfolio")
by_ay <- data_tab("t02_accident_years")
size_dist <- data_tab("t03_claim_sizes")
concentration <- data_tab("t04_concentration")
largest <- data_tab("t05_largest_claims")
delay_dist <- data_tab("t06_delays")
reported <- data_tab("t07_reported_claims")
ibnyr <- data_tab("t08_ibnyr_counts")
status <- data_tab("t09_claim_status")
censoring <- data_tab("t10_censoring")
by_delay <- data_tab("t11_by_reporting_delay")
by_payments <- data_tab("t12_by_number_of_payments")
by_size <- data_tab("t13_by_claim_size")
incremental <- data_tab("t14_incremental_payments")
cumulative <- data_tab("t15_cumulative_payments")
development <- data_tab("t16_development")
calendar <- data_tab("t17_calendar_years")
late <- data_tab("t18_late_payments")
reserves <- data_tab("t19_cl_reserves")

design <- feat_tab("t01_design")
interactions <- feat_tab("t02_interactions")
mix <- feat_tab("t03_mix")
dependent <- feat_tab("t04_dependent_pairs")
cramer <- feat_tab("t05_cramers_v")
combos <- feat_tab("t06_combinations")
top_combos <- feat_tab("t07_top_combinations")
stability <- feat_tab("t08_mix_by_accident_year")
size_level <- feat_tab("t09_claim_size")
sev_effects <- feat_tab("t10_severity_relativities")
injury_age <- feat_tab("t11_injury_by_age")
tail_level <- feat_tab("t12_tail")
delay_level <- feat_tab("t13_delays")
delay_effects <- feat_tab("t14_delay_multipliers")
truncation <- feat_tab("t15_truncation")
payments_band <- feat_tab("t16_payments_by_size")
payments_level <- feat_tab("t17_payments_by_level")
variance <- feat_tab("t18_variance")
payments_r2 <- feat_tab("t19_payments_r2")
pattern <- feat_tab("t20_payment_pattern")
timing <- feat_tab("t21_payment_timing")
exposure <- feat_tab("t22_exposure")
open_level <- feat_tab("t23_open_claims")
os_injury <- feat_tab("t24_outstanding_by_injury")
factors_level <- feat_tab("t25_cl_factors")
cl_level <- feat_tab("t26_cl_reserves")
cl_split <- feat_tab("t27_cl_split")

#########  numbers quoted in the text

p <- setNames(portfolio$value, portfolio$quantity)
cmb <- setNames(combos$value, combos$quantity)
leg <- "Legal Representation"
inj <- "Injury Severity"
age <- "Age of Claimant"
veh <- "Vehicle type"
bus <- "Business use"
## value of column col of a table by level
lv <- function(d, feature, level, col) {
  d[d$feature == feature & d$level == level, col]
}
## share of the variance (%) of an outcome: drop-one of a feature, or the
## additive or saturated model
r2 <- function(outcome, term, col = "drop_one") {
  100 * variance[variance$outcome == outcome & variance$term == term, col]
}
ay <- by_ay[by_ay$AY != "total", ]
ay_total <- by_ay[by_ay$AY == "total", ]
st_total <- status[status$AY == "total", ]
res_total <- reserves[reserves$AY == "total", ]
os_recent <- 100 * sum(ay$outstanding[16:20]) / ay_total$outstanding
os_last <- 100 * ay$outstanding[20] / ay_total$outstanding
cl_error_pct <- 100 * res_total$error / res_total$true_reserves
future <- calendar[!is.na(calendar$outstanding_pct), ]
dev_row <- function(k) development[development$dev == k, ]
n_ref <- nrow(sev_effects)
sev_inside <- sum(sev_effects$lower <= sev_effects$design &
                    sev_effects$design <= sev_effects$upper)
delay_inside <- sum(delay_effects$lower <= delay_effects$design &
                      delay_effects$design <= delay_effects$upper)
inj_high <- exposure[exposure$feature == inj & exposure$level %in% 3:5, ]
```

# What this report covers

This report describes the claims portfolio that `analysis/00_claim-simulation/short_tailed_claims.R` simulates with SynthETIC (Avanzi, Taylor, Wang and Wong 2021), and measures what each of the five claim features does to the claims. It is the SynthETIC counterpart of the description of the Gabrielli and Wuthrich portfolio in `simulated data description.R`: the same kinds of tables and figures, followed by the part that portfolio does not need, the impact of the features.

All numbers come from two scripts in `analysis/00_claim-simulation/`:

- `SynthETIC claims description.R` writes the tables and figures of the portfolio (sections 3 to 7) to `output/tables/00_claim-simulation/data-description/` and `output/figures/00_claim-simulation/data-description/`;
- `SynthETIC feature impact.R` writes those of the features (sections 8 to 10) to the folders `feature-impact/` next to them.

This document reads those tables: every number in the text is computed from them when the document is knitted, so the text cannot differ from the tables. The file name of each table is in the appendix.

## Main findings

1. **Size and maturity.** The portfolio has `r num(p[["claims"]])` claims and `r num(p[["payments"]])` payments over 20 accident years. The nominal ultimate is `r num(p[["nominal ultimate (millions)"]], 1)` million, of which `r num(p[["paid to the valuation date (% of the ultimate)"]], 2)`% is paid at the valuation date. The true outstanding is `r num(p[["true outstanding (millions)"]], 1)` million: `r num(p[["true outstanding, development years 1 to 20 (millions)"]], 1)` million inside the 20 x 20 triangle (the true reserve of the model scripts) and `r num(p[["true outstanding, after development year 20 (millions)"]], 1)` million paid after development year 20.
2. **Where the outstanding sits.** Accident years 16 to 20 hold `r num(os_recent, 1)`% of the true outstanding and accident year 20 alone `r num(os_last, 1)`%. `r num(100 * st_total$os_open / (st_total$os_open + st_total$os_unreported), 1)`% of it is on claims that are reported but not settled, and `r num(100 * st_total$os_unreported / (st_total$os_open + st_total$os_unreported), 1)`% on claims not yet reported.
3. **The chain ladder is close on the whole triangle.** Mack's chain ladder gives a reserve of `r num(res_total$CL_reserves, 1)` million against the true `r num(res_total$true_reserves, 1)` million: an error of `r num(cl_error_pct, 2)`%, which is `r num(abs(res_total$error) / res_total$msep_sqrt, 2)` times its sqrt(msep) of `r num(res_total$msep_sqrt, 1)` million.
4. **The features are simulated as designed.** The numbers of claims in the `r num(cmb[["with a positive designed probability"]])` possible feature combinations fit the designed probabilities (chi-square `r num(cmb[["chi-square of the observed against the designed numbers"]], 1)` on `r num(cmb[["degrees of freedom"]])` degrees of freedom, p-value `r num(cmb[["p-value"]], 2)`). The features are independent of each other except for two designed pairs: Legal Representation with Injury Severity, and Vehicle type with Business use.
5. **Injury Severity is the main feature for every outcome.** Dropping it from the model of all five features removes `r num(r2("log claim size", inj), 1)` percentage points of the `r num(r2("log claim size", "additive", "r2"), 1)`% of the variance of the log claim size that the features explain; Vehicle type follows with `r num(r2("log claim size", veh), 1)` points and Age of Claimant with `r num(r2("log claim size", age), 1)`.
6. **One-way comparisons mislead where the features depend on each other.** The mean claim size of unrepresented claimants is `r num(lv(sev_effects, leg, "N", "ratio_of_means"), 2)` times that of represented ones, although the designed relativity is `r num(lv(sev_effects, leg, "N", "design"), 2)`: unrepresented claims only occur with the lowest injury severity. All else equal, the log-linear model recovers `r num(lv(sev_effects, leg, "N", "estimate"), 2)`. The same happens for Business use Y (`r num(lv(sev_effects, bus, "Y", "ratio_of_means"), 2)` one-way, `r num(lv(sev_effects, bus, "Y", "design"), 2)` designed) and Heavy goods (`r num(lv(sev_effects, veh, "Heavy goods", "ratio_of_means"), 1)` one-way, `r num(lv(sev_effects, veh, "Heavy goods", "design"), 0)` designed).
7. **Injury Severity and the age band 0-15 slow the payments; Vehicle type and Business use do not.** By the end of development year 2, `r num(lv(timing, inj, "1", "dev_2"), 1)`% of the cost of Injury Severity 1 is paid but only `r num(lv(timing, inj, "5", "dev_2"), 1)`% of Injury Severity 5. The pattern of Light Commerical vehicles (`r num(lv(timing, veh, "Light Commerical", "dev_2"), 1)`%) is that of Passenger vehicles (`r num(lv(timing, veh, "Passenger", "dev_2"), 1)`%) although their claims are `r num(lv(sev_effects, veh, "Light Commerical", "design"), 0)` times as large by design.
8. **The features move the reserve far more than the claim count.** Injury Severity 3 to 5 are `r num(sum(inj_high$claims_pct), 1)`% of the claims, `r num(sum(inj_high$ultimate_pct), 1)`% of the nominal ultimate and `r num(sum(inj_high$outstanding_pct), 1)`% of the true outstanding.
9. **A chain ladder per level does not improve the total here.** The sum of the level reserves misses the truth by `r num(min(cl_split$error_pct[-1]), 2)`% to `r num(max(cl_split$error_pct[-1]), 2)`% depending on the feature used for the split, against `r num(cl_split$error_pct[1], 2)`% for the whole triangle. All these differences are well inside one sqrt(msep), and this is one simulation: they are single draws, not biases. The reserves of single levels are much less accurate than the total.

## Conventions

- **True and observed.** The simulation gives every claim to its last payment. "True" means computed on the full simulation; "observed" means known at the valuation date, the end of year 20. A payment is observed when its calendar year is 20 or earlier. Unless a table says "observed", it is true.
- **Development year.** Development year = calendar year of the payment - accident year + 1, so development year 1 is the accident year. The triangles have 20 development years; the payments after development year 20 are in no triangle and are shown apart (`r tab_ref("late")`), as in the Mack, ODP GLM, bCCNN and NN chain-ladder scripts. The file `triangle.csv` of the simulation folds them into development year 20 and is not used here.
- **Nominal and constant values.** The claim size and the payment sizes of SynthETIC are in constant values (money of time 0). Every payment is then inflated by 2% a year from time 0 to its payment date: these nominal payments fill the triangles. Claim-size tables are in constant values; triangles, ultimates, outstanding amounts and reserves are nominal.
- **Reference levels.** Relativities and multipliers are given against the reference level of each feature: Legal Representation Y, Injury Severity 2, Age of Claimant 30-50, Vehicle type Passenger and Business use N (the levels with delay multiplier 1).
- **Level names.** The level names are those of the data. "Light Commerical" is spelt this way in the data and is left as it is.

# The simulation

`r tab_ref("simulation")` lists the settings of `short_tailed_claims.R`. The delays are those of environment 1 of Al-Mudafer, Avanzi, Taylor and Wong (2021): their means do not depend on the claim size. The features are an addition: three come from SynthETIC's test covariates (Legal Representation, Injury Severity, Age of Claimant) and two are added by the script (Vehicle type, Business use), together with the multipliers of the two delays.

```{r simulation}
simulation <- data.frame(
  step = c("Claim numbers", "Claim size before the features", "Features",
           "Notification delay", "Settlement delay", "Number of payments",
           "Payment sizes and times", "Inflation", "Valuation"),
  setting = c(
    paste("Poisson with mean 200,000 x 0.1 = 20,000 per accident year, 20",
          "accident years; the occurrence is uniform within the year."),
    paste("The fifth power of a normal variable with mean 9.5 and standard",
          "deviation 3, truncated to sizes of 30 or more; the same for",
          "every accident year."),
    paste("Five features, 2 x 6 x 5 x 4 x 2 = 480 combinations of levels.",
          "A claim draws its combination with a probability proportional",
          "to the product of the frequency relativities; its size is",
          "multiplied by the product of the severity relativities, and all",
          "sizes are rescaled so that the total claim cost is unchanged.",
          "The frequency relativities change the mix of the claims, not",
          "their number."),
    paste("Weibull with mean 0.49305517 / 4 = 0.1233 years and coefficient",
          "of variation 1.919, times the notification multipliers of the",
          "claim's five levels."),
    paste("Weibull with mean 6.575582 / 4 = 1.6439 years and coefficient of",
          "variation 1.023, measured from the notification, times the",
          "settlement multipliers of the claim's five levels."),
    paste("Drawn from the claim size only: 1 or 2 payments up to 7,500;",
          "2 or 3 up to 15,000; above that 4 plus a geometric number, with",
          "mean min(8, 4 + ln(size / 15,000))."),
    paste("SynthETIC's default split of the claim into payments. All",
          "payments lie between the notification and the settlement; the",
          "last payment is at the settlement."),
    paste("2% a year on every payment from time 0 to its payment date; no",
          "superimposed inflation."),
    paste("End of year 20: 20 accident years and 20 development years",
          "(annual).")
  )
)
kable(simulation, col.names = c("Step", "Setting"),
      caption = tab_cap("simulation"))
```

# The portfolio

## Size

`r tab_ref("portfolio")` gives the portfolio at the valuation date. `r num(p[["claims closed"]])` claims are closed, `r num(p[["claims open"]])` are open and `r num(p[["claims not yet reported"]])` are not yet reported. Of the `r num(p[["payments"]])` payments, `r num(p[["payments after the valuation date"]])` fall after the valuation date. They are the true outstanding of `r num(p[["true outstanding (millions)"]], 1)` million, which is `r num(100 - p[["paid to the valuation date (% of the ultimate)"]], 2)`% of the nominal ultimate.

```{r portfolio}
kable(data.frame(portfolio$quantity, smart(portfolio$value)),
      col.names = c("Quantity", "Value"), align = "lr",
      caption = tab_cap("portfolio"))
```

The claim cost in constant values is `r num(p[["claim cost in constant values (millions)"]], 1)` million and the nominal ultimate `r num(p[["nominal ultimate (millions)"]], 1)` million: inflation adds `r num(100 * (p[["nominal ultimate (millions)"]] / p[["claim cost in constant values (millions)"]] - 1), 1)`% over the whole portfolio, because it runs from time 0 and the average payment is made well into the 20 years.

## Accident years

`r tab_ref("accident_years")` and `r fig_ref("ay")` show the accident years. The number of claims varies between `r num(min(ay$claims))` and `r num(max(ay$claims))` around the designed 20,000, and the claim cost in constant values has no trend. The nominal ultimate rises from `r num(ay$ultimate[1], 0)` million in accident year 1 to `r num(ay$ultimate[20], 0)` million in accident year 20: this growth is inflation only (the ratio of nominal to constant values rises from `r num(ay$nominal_to_constant[1], 3)` to `r num(ay$nominal_to_constant[20], 3)`).

The mean claim size of accident year `r ay$AY[which.max(ay$mean_size)]` (`r num(max(ay$mean_size))`) stands out. It is one claim of `r num(max(ay$max_size), 1)` million (`r tab_ref("largest")`); the median is unaffected. Means by accident year or by level are sensitive to such claims, so the tables give medians next to them.

```{r accident-years}
show(by_ay, "accident_years",
     c("AY", "Claims", "Claim cost", "Mean size", "Median size",
       "Largest claim", "Nominal ultimate", "Paid", "Outstanding", "Paid %",
       "Nominal / constant"),
     digits = c(0, 0, 1, 0, 0, 1, 1, 1, 1, 1, 3))
```

```{r fig-ay, out.width = "6.3in", fig.cap = fig_cap("ay")}
include_graphics(data_fig("01 accident years"))
```

# Claim sizes and delays

## Claim sizes

`r tab_ref("claim_sizes")` gives the distribution of the claim size before the features (raw), after them (adjusted) and of the nominal ultimate of a claim. The features leave the mean unchanged by construction (`r num(size_dist$raw[size_dist$statistic == "mean"])`) but lower the median from `r num(size_dist$raw[size_dist$statistic == "median"])` to `r num(size_dist$adjusted[size_dist$statistic == "median"])` and lengthen the tail: the coefficient of variation rises from `r num(size_dist$raw[size_dist$statistic == "coefficient of variation"], 2)` to `r num(size_dist$adjusted[size_dist$statistic == "coefficient of variation"], 2)` and the largest claim from `r num(size_dist$raw[size_dist$statistic == "maximum"] / 1e6, 1)` million to `r num(size_dist$adjusted[size_dist$statistic == "maximum"] / 1e6, 1)` million (`r fig_ref("size")`).

```{r claim-sizes}
kable(data.frame(size_dist$statistic, smart(size_dist$raw),
                 smart(size_dist$adjusted), smart(size_dist$nominal)),
      col.names = c("Statistic", "Before the features", "After the features",
                    "Nominal ultimate"),
      align = "lrrr", caption = tab_cap("claim_sizes"))
```

```{r fig-size, out.width = "4.72in", fig.cap = fig_cap("size")}
include_graphics(data_fig("02 claim size distribution"))
```

The cost is concentrated in few claims (`r tab_ref("concentration")`, `r fig_ref("lorenz")`). The largest 1% of the claims hold `r num(concentration$adjusted[2], 1)`% of the cost, against `r num(concentration$raw[2], 1)`% before the features, and the Gini coefficient rises from `r num(concentration$raw[6], 3)` to `r num(concentration$adjusted[6], 3)`.

```{r concentration}
kable(data.frame(concentration$statistic, smart(concentration$raw),
                 smart(concentration$adjusted)),
      col.names = c("Statistic", "Before the features", "After the features"),
      align = "lrr", caption = tab_cap("concentration"))
```

```{r fig-lorenz, out.width = "4.72in", fig.cap = fig_cap("lorenz")}
include_graphics(data_fig("03 Lorenz curves"))
```

The ten largest claims (`r tab_ref("largest")`) are not the claims with the largest size before the features: they are made by the product of the severity relativities. Among the ten, `r sum(largest[["Vehicle type"]] != "Passenger")` are commercial vehicles (not Passenger) and `r sum(largest[["Injury Severity"]] %in% 4:5)` have Injury Severity 4 or 5. All ten are closed at the valuation date, so they are in the observed triangle and not in the outstanding.

```{r largest}
largest_tab <- data.frame(largest$claim_no, largest$AY,
                          largest$claim_size / 1e6, largest$ultimate / 1e6,
                          largest$claim_size_raw / 1e6,
                          largest$sev_relativity, largest$setldel,
                          largest$no_payment,
                          largest[, c(leg, inj, age, veh, bus)])
show(largest_tab, "largest",
     c("Claim", "AY", "Claim size", "Nominal ultimate",
       "Size before the features", "Size factor", "Settlement delay",
       "Payments", "Legal", "Injury", "Age", "Vehicle", "Business"),
     digits = c(0, 0, 1, 1, 2, 2, 2, 0, 0, 0, 0, 0, 0))
```

The size factor in `r tab_ref("largest")` is the claim size over the size before the features: the product of the severity relativities of the claim's levels times the rescaling constant (section 8).

## Delays

`r tab_ref("delays")` and `r fig_ref("delays")` describe the two delays. Reporting is fast: the mean notification delay is `r num(delay_dist$notification[1], 3)` years and the median `r num(delay_dist$notification[delay_dist$statistic == "median"], 3)` years; `r num(delay_dist$notification[delay_dist$statistic == "% over 1 year"], 1)`% of the claims take more than a year. The settlement delay, measured from the notification, has a mean of `r num(delay_dist$settlement[1], 2)` years and a median of `r num(delay_dist$settlement[delay_dist$statistic == "median"], 2)` years; `r num(delay_dist$settlement[delay_dist$statistic == "% over 5 years"], 1)`% of the claims take more than five years and the longest takes `r num(delay_dist$settlement[delay_dist$statistic == "maximum"], 1)` years.

```{r delays}
kable(data.frame(delay_dist$statistic, smart(delay_dist$notification),
                 smart(delay_dist$settlement),
                 smart(delay_dist$occurrence_to_settlement)),
      col.names = c("Statistic", "Notification delay", "Settlement delay",
                    "Occurrence to settlement"),
      align = "lrrr", caption = tab_cap("delays"))
```

```{r fig-delays, out.width = "4.72in", fig.cap = fig_cap("delays")}
include_graphics(data_fig("04 delays"))
```

# Reporting and claim status

## Reported claims

`r tab_ref("reported")` is the triangle of the claim numbers by reporting delay (the analogue of Table 3 of Gabrielli and Wuthrich 2018). `r num(100 * by_delay$claims[1] / sum(by_delay$claims), 1)`% of the claims are reported in their accident year and `r num(100 * by_delay$claims[2] / sum(by_delay$claims), 1)`% in the next year; the longest reporting delay is `r max(by_delay$rep_delay)` years. `r num(st_total$unreported)` claims are not yet reported at the valuation date, `r num(status$unreported[status$AY == "20"])` of them in accident year 20.

```{r reported}
show(reported, "reported",
     c("AY", paste0("T = ", 0:(ncol(reported) - 4)), "Reported",
       "Not yet reported"),
     digits = 0)
```

The chain ladder on this triangle (`r tab_ref("ibnyr")`) predicts `r num(ibnyr$CL_IBNYR[ibnyr$AY == "total"], 1)` unreported claims with a sqrt(msep) of `r num(ibnyr$msep_sqrt[ibnyr$AY == "total"], 1)`, against the true `r num(ibnyr$true_IBNYR[ibnyr$AY == "total"])`: the reporting pattern is stable and the number of unreported claims is easy to estimate.

```{r ibnyr}
show(ibnyr, "ibnyr",
     c("AY", "Reported", "Chain-ladder estimate", "sqrt(msep)",
       "True number"),
     digits = c(0, 0, 1, 1, 0))
```

## Status at the valuation date

`r tab_ref("status")` and `r fig_ref("open")` split the claims by status. The open claims hold `r num(st_total$os_open, 1)` million of the true outstanding and the unreported claims `r num(st_total$os_unreported, 1)` million. `r num(st_total$open_no_payment)` of the open claims have no payment yet at the valuation date (`r num(status$open_no_payment[status$AY == "20"])` of them in accident year 20). For a model that uses the features this matters: the features of the open claims are known at the valuation date, those of the unreported claims are not.

```{r status}
show(status, "status",
     c("AY", "Claims", "Closed", "Open", "Not yet reported",
       "Open, no payment yet", "Outstanding, open",
       "Outstanding, not yet reported"),
     digits = c(0, 0, 0, 0, 0, 0, 1, 1))
```

```{r fig-open, out.width = "6.3in", fig.cap = fig_cap("open")}
include_graphics(data_fig("09 open and unreported claims"))
```

The valuation date also selects the claims: the closed claims are the fast ones (`r tab_ref("censoring")`). Over all accident years the closed claims have a mean settlement delay of `r num(censoring$mean_setldel[1], 2)` years and the open claims of `r num(censoring$mean_setldel[2], 2)` years; in accident year 20 the closed claims have `r num(censoring$mean_setldel[4], 2)` years. Statistics of the closed claims therefore understate the delays (and the sizes), which is why this report measures the effects of the features on the full simulation.

```{r censoring}
show(censoring, "censoring",
     c("Accident years", "Status", "Claims", "Mean size", "Mean ultimate",
       "Mean notification delay", "Mean settlement delay",
       "Mean number of payments"),
     digits = c(0, 0, 0, 0, 0, 3, 3, 2))
```

## Reporting delay, number of payments and claim size

`r tab_ref("by_delay")` to `r tab_ref("by_size")` and `r fig_ref("bars_delay")` and `r fig_ref("bars_payments")` are the analogues of Figures A2 and A3 of Gabrielli and Wuthrich (2018). The claim size and the settlement delay do not depend on the reporting delay (the rows from T = 3 on rest on few claims). The number of payments rises with the claim size, by design: claims of 1 to 3 payments are small and together hold `r num(sum(by_payments$cost_pct[1:3]), 1)`% of the cost. SynthETIC has no recoveries, so panel (c) of Figure A3 is replaced by the settlement delay.

```{r by-delay}
show(by_delay, "by_delay",
     c("T", "Claims", "Mean size", "Mean number of payments",
       "Mean settlement delay"),
     digits = c(0, 0, 0, 2, 2))
```

```{r fig-bars-delay, out.width = "6.3in", fig.cap = fig_cap("bars_delay")}
include_graphics(data_fig("11 by reporting delay"))
```

```{r by-payments}
by_payments$no_payment <- ifelse(by_payments$no_payment == 15, "15 or more",
                                 by_payments$no_payment)
show(by_payments, "by_payments",
     c("Payments", "Claims", "Mean size", "Median size",
       "Mean settlement delay", "% of the cost"),
     digits = c(0, 0, 0, 0, 2, 2), align = "rrrrrr")
```

```{r fig-bars-payments, out.width = "6.3in", fig.cap = fig_cap("bars_payments")}
include_graphics(data_fig("12 by number of payments"))
```

`r tab_ref("by_size")` shows the claims by size band. The settlement delay rises with the claim size, from `r num(by_size$mean_setldel[1], 2)` years in the lowest band to `r num(by_size$mean_setldel[nrow(by_size)], 2)` years in the highest. The generator has no such link (the delays of environment 1 do not depend on the size): it is created by the features, which make some claims both larger and slower (section 9). The claims over 500,000 are `r num(sum(by_size$claims_pct[6:7]), 1)`% of the claims and hold `r num(sum(by_size$outstanding_pct[6:7]), 1)`% of the true outstanding.

```{r by-size}
show(by_size, "by_size",
     c("Claim size", "Claims", "% of claims", "% of cost",
       "Mean notification delay", "Mean settlement delay",
       "Mean number of payments", "% of outstanding", "Paid %"),
     digits = c(0, 0, 1, 1, 3, 2, 2, 1, 1))
```

# Triangles and development

## The triangles

`r tab_ref("incremental")` and `r tab_ref("cumulative")` are the observed triangles of the nominal payments, the data of the Mack, ODP GLM and bCCNN scripts. The last column of `r tab_ref("cumulative")` is the true ultimate at development year 20. `r fig_ref("triangle")` shows the whole 20 x 20 square, with the future cells below the staircase.

```{r incremental}
show(incremental, "incremental", c("AY", 1:20), digits = 1)
```

```{r cumulative}
show(cumulative, "cumulative", c("AY", 1:20, "True ultimate"), digits = 1)
```

```{r fig-triangle, out.width = "4.72in", fig.cap = fig_cap("triangle")}
include_graphics(data_fig("06 triangle"))
```

## Development factors and payment pattern

`r tab_ref("development")` gives the chain-ladder factors of the observed triangle next to those of the full simulation, the range of the observed individual factors, and the share of the ultimate paid by development year. The business is short-tailed in the sense of environment 1: `r num(dev_row(1)$paid_pct_true, 1)`% of the ultimate is paid in the accident year, `r num(dev_row(2)$paid_pct_true, 1)`% by the end of development year 2, `r num(dev_row(5)$paid_pct_true, 1)`% by year 5 and `r num(dev_row(10)$paid_pct_true, 1)`% by year 10. The factors of the observed triangle are close to those of the full simulation (the first factor is `r num(dev_row(1)$cl_factor, 4)` against `r num(dev_row(1)$cl_factor_true, 4)`).

```{r development}
show(development, "development",
     c("Development year j", "Factor j to j + 1, observed",
       "Factor, full simulation", "Smallest individual factor",
       "Largest individual factor", "Paid %, chain ladder", "Paid %, true",
       "Paid in the year %, true"),
     digits = c(0, 4, 4, 4, 4, 2, 2, 2))
```

`r fig_ref("dev")` shows that the accident years develop alike, and `r fig_ref("link")` the individual factors behind the first six chain-ladder factors: they scatter around the chain-ladder factor without a trend over the accident years. The largest first factor (`r num(dev_row(1)$link_max, 3)`) belongs to accident year 3, whose largest claim is paid almost entirely in development year 2.

```{r fig-dev, out.width = "6.3in", fig.cap = fig_cap("dev")}
include_graphics(data_fig("05 development by accident year"))
```

```{r fig-link, out.width = "6.3in", fig.cap = fig_cap("link")}
include_graphics(data_fig("07 development factors"))
```

## Calendar years and inflation

`r tab_ref("calendar")` and `r fig_ref("calendar")` give the payments by calendar year. The index is the nominal amount over the amount in constant values: it rises by 2% a year. After the valuation date only the 20 simulated accident years run off: `r num(future$outstanding_pct[1], 1)`% of the true outstanding is paid in the first year after the valuation date and `r num(sum(future$outstanding_pct[1:5]), 1)`% within five years.

```{r calendar}
show(calendar, "calendar",
     c("Calendar year", "Payments", "Nominal", "Constant values", "Index",
       "% of the outstanding"),
     digits = c(0, 0, 1, 1, 3, 1), align = "rrrrrr")
```

```{r fig-calendar, out.width = "4.72in", fig.cap = fig_cap("calendar")}
include_graphics(data_fig("08 calendar years"))
```

The payments after development year 20 (`r tab_ref("late")`) are `r num(sum(late$amount), 1)` million on `r num(sum(late$claims))` claims, `r num(100 * sum(late$amount) / ay_total$ultimate, 3)`% of the nominal ultimate. They are in no triangle and no model predicts them.

```{r late}
show(late, "late",
     c("AY", "Claims", "Payments", "Amount", "% of the ultimate of the AY"),
     digits = c(0, 0, 0, 2, 3))
```

# Chain-ladder reserves

`r tab_ref("cl_reserves")` and `r fig_ref("reserves")` compare Mack's chain-ladder reserves with the true outstanding of the triangle (the analogue of Table 7 of Gabrielli and Wuthrich 2018; the same numbers as `output/tables/01_Mack/`). The total reserve is `r num(res_total$CL_reserves, 1)` million against `r num(res_total$true_reserves, 1)` million, an error of `r num(res_total$error, 1)` million. The largest error is in accident year 20 (`r num(reserves$error[reserves$AY == "20"], 1)` million, `r num(abs(reserves$error[reserves$AY == "20"]) / reserves$msep_sqrt[reserves$AY == "20"], 2)` times its sqrt(msep)).

```{r cl-reserves}
show(reserves, "cl_reserves",
     c("AY", "Paid", "Chain-ladder reserve", "sqrt(msep)",
       "True outstanding", "Error"),
     digits = 1)
```

```{r fig-reserves, out.width = "6.3in", fig.cap = fig_cap("reserves")}
include_graphics(data_fig("10 chain-ladder reserves"))
```

# The features: design and mix

## Design

`r tab_ref("f_design")` gives the design of the five features. The probability of a level is the sum of the designed probabilities of the combinations that contain it. The severity relativity is the factor on the claim size, all else equal; the column next to it is the same relative to the reference level. The two multipliers scale the mean (and every quantile) of the notification and of the settlement delay.

```{r f-design}
design_tab <- data.frame(design$feature, design$level,
                         100 * design$design_prob, design$freq_relativity,
                         design$sev_relativity, design$sev_vs_reference,
                         design$notidel_mult, design$setldel_mult,
                         design$rationale)
show(design_tab, "f_design",
     c("Feature", "Level", "Probability %", "Frequency relativity",
       "Severity relativity", "Severity against reference",
       "Notification multiplier", "Settlement multiplier",
       "Reason for the delay multipliers"),
     digits = c(0, 0, 2, 3, 2, 3, 2, 2, 0))
```

Three pairs of features have relativities that are not 1 (`r tab_ref("f_interactions")`). In the frequency relativities, Legal Representation N is only possible with Injury Severity 1, and Vehicle type depends on Business use (Heavy goods is not possible with Business use N). In the severity relativities, Injury Severity 4 and 5 are reduced for the two oldest age bands. Of the 480 combinations, `r num(cmb[["with a positive designed probability"]])` are possible.

The severity relativities are rescaled so that the total claim cost stays the same: a claim's size is its size before the features times the product of the severity relativities of its levels times one constant, which is `r num(cmb[["rescaling constant of the severity relativities"]], 4)` in this simulation (`SynthETIC feature impact.R` checks this identity for every claim). Relativities are therefore only meaningful as ratios between levels.

```{r f-interactions}
show(interactions, "f_interactions",
     c("Relativity of", "Feature 1", "Feature 2", "Level 1", "Level 2",
       "Relativity"),
     digits = 2)
```

## The portfolio mix

`r tab_ref("f_mix")` and `r fig_ref("f_mix")` compare the observed share of the claims with the designed probability. The standardised differences z are all below `r num(ceiling(10 * max(abs(mix$z))) / 10, 1)` in absolute value. Three levels are small: Heavy goods has `r num(lv(mix, veh, "Heavy goods", "n"))` claims, Medium goods `r num(lv(mix, veh, "Medium goods", "n"))` and Injury Severity 5 `r num(lv(mix, inj, "5", "n"))`. Their averages are noisy and should not be over-read.

```{r f-mix}
show(mix, "f_mix",
     c("Feature", "Level", "Claims", "Observed %", "Designed %", "z"),
     digits = c(0, 0, 0, 3, 3, 2))
```

```{r fig-f-mix, out.width = "4.72in", fig.cap = fig_cap("f_mix")}
include_graphics(feat_fig("01 portfolio mix"))
```

`r tab_ref("f_dependent")` and `r fig_ref("f_pairs")` show the two dependent pairs. All `r num(lv(mix, leg, "N", "n"))` unrepresented claims have Injury Severity 1 (they are `r num(dependent$within_level_1_pct[dependent$level_1 == "1" & dependent$level_2 == "N"], 1)`% of that level). The share of Business use Y is `r num(dependent$within_level_1_pct[dependent$level_1 == "Passenger" & dependent$level_2 == "Y"], 1)`% for Passenger vehicles, `r num(dependent$within_level_1_pct[dependent$level_1 == "Light Commerical" & dependent$level_2 == "Y"], 1)`% for Light Commerical, `r num(dependent$within_level_1_pct[dependent$level_1 == "Medium goods" & dependent$level_2 == "Y"], 1)`% for Medium goods and 100% for Heavy goods. These two dependencies are the source of all confounding in the one-way tables that follow.

```{r f-dependent}
show(dependent, "f_dependent",
     c("Pair", "Level of the first", "Level of the second", "Claims",
       "Observed %", "Designed %", "% within the level of the first"),
     digits = c(0, 0, 0, 0, 3, 3, 1))
```

```{r fig-f-pairs, out.width = "4.72in", fig.cap = fig_cap("f_pairs")}
include_graphics(feat_fig("02 dependent features"))
```

Cramer's V (`r tab_ref("f_cramer")`) is `r num(cramer[cramer$feature == leg, inj], 3)` for Legal Representation with Injury Severity and `r num(cramer[cramer$feature == veh, bus], 3)` for Vehicle type with Business use, and at most `r num(max(as.matrix(cramer[, -1])[as.matrix(cramer[, -1]) < 0.1]), 4)` for every other pair, including each feature with the accident year. The V of Legal Representation with Injury Severity looks modest although the dependence is total for unrepresented claims, because they are only `r num(lv(mix, leg, "N", "share_pct"), 1)`% of the claims.

```{r f-cramer}
show(cramer, "f_cramer",
     c("", "Legal Representation", "Injury Severity", "Age of Claimant",
       "Vehicle type", "Business use", "Accident year"),
     digits = 4)
```

## Combinations of levels

`r num(cmb[["observed"]])` of the `r num(cmb[["with a positive designed probability"]])` possible combinations occur (`r tab_ref("f_combos")`); those that do not occur have an expected number of claims of at most `r num(cmb[["largest expected number of claims of those not observed"]], 1)`. The feature space is thin: `r num(cmb[["observed with fewer than 100 claims"]])` of the observed combinations have fewer than 100 claims, and the `r num(cmb[["cells accident year x combination with claims"]])` cells of accident year and combination, the units of the NN chain ladder, hold a median of `r num(cmb[["median number of claims per cell"]])` claims.

```{r f-combos}
kable(data.frame(combos$quantity, smart(combos$value)),
      col.names = c("Quantity", "Value"), align = "lr",
      caption = tab_cap("f_combos"))
```

The ten combinations with the largest cost (`r tab_ref("f_top")`) are all ordinary profiles: Legal Representation Y, Passenger, Business use N. Together they hold `r num(top_combos$cum_cost_pct[10], 1)`% of the cost and `r num(sum(top_combos$claims_pct), 1)`% of the claims.

```{r f-top}
show(top_combos, "f_top",
     c("Legal", "Injury", "Age", "Vehicle", "Business", "Claims",
       "% of claims", "% of cost", "Cumulative % of cost",
       "Designed severity against reference", "Median size"),
     digits = c(0, 0, 0, 0, 0, 0, 2, 2, 2, 3, 0))
```

## Stability over the accident years

The mix of the levels does not drift (`r tab_ref("f_stability")`, `r fig_ref("f_mix_ay")`): the p-values of the chi-square tests of level against accident year are between `r num(min(stability$chi2_p_value), 2)` and `r num(max(stability$chi2_p_value), 2)`. This is why the features do not bias the chain ladder of the whole triangle. The shares of the cost are less stable than the shares of the claims, because few large claims move them: the share of Heavy goods in the nominal ultimate of an accident year ranges from `r num(lv(stability, veh, "Heavy goods", "ultimate_pct_min"), 1)`% to `r num(lv(stability, veh, "Heavy goods", "ultimate_pct_max"), 1)`%.

```{r f-stability}
show(stability, "f_stability",
     c("Feature", "Level", "Fewest claims in an AY", "% of claims, lowest",
       "% of claims, mean", "% of claims, highest", "% of ultimate, lowest",
       "% of ultimate, mean", "% of ultimate, highest", "Chi-square p-value"),
     digits = c(0, 0, 0, 2, 2, 2, 2, 2, 2, 3))
```

```{r fig-f-mix-ay, out.width = "6.3in", fig.cap = fig_cap("f_mix_ay")}
include_graphics(feat_fig("13 mix by accident year"))
```

# Impact of the features on the claims

## Claim size

`r tab_ref("f_size")` and `r fig_ref("f_size")` give the claim size by level. A level moves the whole distribution without changing its shape: the quartile range has about the same width on the log scale in every level. The ratio of the share of the cost to the share of the claims runs from `r num(min(size_level$cost_to_claims), 2)` (`r size_level$feature[which.min(size_level$cost_to_claims)]` `r size_level$level[which.min(size_level$cost_to_claims)]`) to `r num(max(size_level$cost_to_claims), 1)` (`r size_level$feature[which.max(size_level$cost_to_claims)]` `r size_level$level[which.max(size_level$cost_to_claims)]`). Injury Severity 6 has the lowest relativity of its feature: the feature is a code, not a scale on which size rises.

```{r f-size}
show(size_level, "f_size",
     c("Feature", "Level", "Claims", "% of claims", "% of cost",
       "Cost / claims", "Mean", "Median", "Geometric mean", "25%", "75%",
       "95%", "99%", "CV", "Standard error of the mean %"),
     digits = c(0, 0, 0, 2, 2, 2, 0, 0, 0, 0, 0, 0, 0, 2, 1))
```

```{r fig-f-size, out.width = "4.72in", fig.cap = fig_cap("f_size")}
include_graphics(feat_fig("04 claim size by level"))
```

`r tab_ref("f_sev")` and `r fig_ref("f_sev")` are the central table and figure on the claim size. For every level other than the reference level they compare three numbers: the designed relativity; the one-way ratio (the mean, and the geometric mean, of the level over that of the reference level); and the all-else-equal estimate, the exponential of the coefficient of the linear model of the log claim size on the five features, with its 95% interval. That model is the structure of the simulator itself: the log claim size is the log size before the features plus the sum of the log relativities.

The all-else-equal estimates recover the design: `r sev_inside` of the `r n_ref` intervals contain the designed relativity. The one-way ratios do not where the features depend on each other: Legal Representation N (`r num(lv(sev_effects, leg, "N", "ratio_of_means"), 3)` against `r num(lv(sev_effects, leg, "N", "design"), 2)`), Business use Y (`r num(lv(sev_effects, bus, "Y", "ratio_of_means"), 3)` against `r num(lv(sev_effects, bus, "Y", "design"), 2)`) and the three commercial vehicle types. For Heavy goods the one-way ratio of the means (`r num(lv(sev_effects, veh, "Heavy goods", "ratio_of_means"), 1)`) also carries the largest claim of the portfolio.

```{r f-sev}
show(sev_effects, "f_sev",
     c("Feature", "Level", "Designed", "One-way, ratio of means",
       "One-way, ratio of geometric means", "All else equal",
       "95% interval, lower", "95% interval, upper"),
     digits = c(0, 0, 3, 3, 3, 3, 3, 3))
```

```{r fig-f-sev, out.width = "4.72in", fig.cap = fig_cap("f_sev")}
include_graphics(feat_fig("05 severity relativities"))
```

`r tab_ref("f_injury_age")` and `r fig_ref("f_injury_age")` look at the two features with a designed severity interaction. The median claim size runs from `r num(min(injury_age$median_size))` to `r num(max(injury_age$median_size))` over the 30 cells. The column "against main effects" is the geometric mean of the cell over the fit of the two main effects; the designed interaction (0.90 to 0.97 in four cells) is of the same size as the noise of the cells of Injury Severity 5 and 6, which have `r num(min(injury_age$n[injury_age$injury %in% 5:6]))` to `r num(max(injury_age$n[injury_age$injury %in% 5:6]))` claims each. The interaction is too small to be seen in the data.

```{r f-injury-age}
show(injury_age, "f_injury_age",
     c("Injury Severity", "Age of Claimant", "Claims", "Median size",
       "Against main effects", "Designed interaction"),
     digits = c(0, 0, 0, 0, 3, 2), align = "rrrrrr")
```

```{r fig-f-injury-age, out.width = "4.72in", fig.cap = fig_cap("f_injury_age")}
include_graphics(feat_fig("08 injury by age"))
```

The features decide who is in the tail (`r tab_ref("f_tail")`). Injury Severity 3 to 5 are `r num(sum(tail_level$claims_pct[tail_level$feature == inj & tail_level$level %in% 3:5]), 1)`% of the claims but `r num(sum(tail_level$top_1pct_pct[tail_level$feature == inj & tail_level$level %in% 3:5]), 1)`% of the largest 1% of the claims; `r num(lv(tail_level, veh, "Heavy goods", "over_1m_pct"), 1)`% of the Heavy goods claims exceed 1 million against `r num(lv(tail_level, veh, "Passenger", "over_1m_pct"), 1)`% of the Passenger claims.

```{r f-tail}
show(tail_level[, c("feature", "level", "claims_pct", "top_1pct_pct", "lift",
                    "over_1m_pct")],
     "f_tail",
     c("Feature", "Level", "% of all claims", "% of the largest 1%",
       "Ratio of the two", "% of its claims over 1 million"),
     digits = c(0, 0, 2, 2, 2, 2))
```

## Delays

`r tab_ref("f_delays")` gives the two delays by level. The mean multiplier of a level's claims is not the level's own multiplier, because the multipliers of the other four features are mixed in: for Legal Representation N it is `r num(lv(delay_level, leg, "N", "mean_setldel_mult"), 3)` for the settlement delay, not the designed `r num(lv(design, leg, "N", "setldel_mult"), 2)`, since these claims all have Injury Severity 1 (multiplier `r num(lv(design, inj, "1", "setldel_mult"), 2)`). The base columns divide the mean delay by the mean multiplier: they are close to the means before the features (0.1233 and 1.6439 years) in every level, which confirms that the multipliers act as designed.

```{r f-delays}
show(delay_level, "f_delays",
     c("Feature", "Level", "Notification: mean", "median", "mean multiplier",
       "% reported in the AY", "base", "Settlement: mean", "median", "90%",
       "mean multiplier", "% over 5 years", "base"),
     digits = c(0, 0, 3, 3, 3, 1, 3, 2, 2, 2, 3, 1, 3))
```

`r tab_ref("f_delay_mult")` and `r fig_ref("f_delay_mult")` compare the designed multipliers with the one-way ratios of the means and with the all-else-equal estimates of the log-linear model (a multiplier scales the Weibull scale, so the log delay is a constant plus the sum of the log multipliers plus an error with the same distribution for every claim). `r delay_inside` of the `r nrow(delay_effects)` intervals contain the designed multiplier. The settlement delay is moved by Injury Severity (`r num(lv(design, inj, "1", "setldel_mult"), 2)` to `r num(lv(design, inj, "5", "setldel_mult"), 2)`), by the age band 0-15 (`r num(lv(design, age, "0-15", "setldel_mult"), 2)`) and by Legal Representation N (`r num(lv(design, leg, "N", "setldel_mult"), 2)`); the multipliers of Vehicle type and Business use are close to 1. The notification multipliers are between `r num(min(design$notidel_mult), 2)` and `r num(max(design$notidel_mult), 2)`, but they act on a delay of about a month and a half: the share of the claims reported in the accident year only ranges from `r num(min(delay_level$same_year_pct), 1)`% to `r num(max(delay_level$same_year_pct), 1)`% over the 19 levels.

```{r f-delay-mult}
show(delay_effects, "f_delay_mult",
     c("Delay", "Feature", "Level", "Designed", "One-way, ratio of means",
       "All else equal", "95% interval, lower", "95% interval, upper"),
     digits = c(0, 0, 0, 2, 3, 3, 3, 3))
```

```{r fig-f-delay-mult, out.width = "6.3in", fig.cap = fig_cap("f_delay_mult")}
include_graphics(feat_fig("06 delay multipliers"))
```

These effects cannot be read off the claims that are closed at the valuation date (`r tab_ref("f_truncation")`). On the closed claims the estimate for Injury Severity 5 is `r num(lv(truncation, inj, "5", "closed"), 2)` and on the closed claims of the last five accident years `r num(lv(truncation, inj, "5", "closed_last_5_years"), 2)`, against the designed `r num(lv(truncation, inj, "5", "design"), 2)`: the slow claims of a slow level are still open, so the closed ones look alike in every level.

```{r f-truncation}
show(truncation, "f_truncation",
     c("Feature", "Level", "Designed", "All claims", "Closed claims",
       "Closed claims, accident years 16 to 20"),
     digits = c(0, 0, 2, 3, 3, 3))
```

## Number of payments

The features have no direct effect on the number of payments: SynthETIC draws it from the claim size (`r tab_ref("f_pay_band")`). A level with larger claims therefore has more payments (`r tab_ref("f_pay_level")`): the mean runs from `r num(min(payments_level$mean_no_payment), 2)` to `r num(max(payments_level$mean_no_payment), 2)`, and in every level it agrees with the mean of the designed number given the claim sizes of its claims. `r fig_ref("f_np")` shows the same: at a given claim size all levels of Injury Severity lie on the designed curve.

```{r f-pay-band}
show(payments_band, "f_pay_band",
     c("Claim size", "Claims", "% of claims", "Mean number of payments",
       "Designed mean"),
     digits = c(0, 0, 1, 3, 3))
```

```{r f-pay-level}
show(payments_level, "f_pay_level",
     c("Feature", "Level", "% up to 7,500", "% 7,500 to 15,000",
       "% at the cap of 8", "Mean number of payments",
       "Designed mean given the sizes"),
     digits = c(0, 0, 1, 1, 1, 2, 2))
```

```{r fig-f-np, out.width = "4.72in", fig.cap = fig_cap("f_np")}
include_graphics(feat_fig("11 number of payments"))
```

`r tab_ref("f_pay_r2")` puts this in one number. The features explain `r num(100 * payments_r2$r2[1], 1)`% of the variance of the number of payments, the designed mean given the claim size `r num(100 * payments_r2$r2[2], 1)`%, and both together `r num(100 * payments_r2$r2[3], 1)`%: once the size is known the features add nothing.

```{r f-pay-r2}
payments_r2$r2 <- 100 * payments_r2$r2
show(payments_r2, "f_pay_r2", c("Model", "Variance explained %"), digits = 2)
```

## Variance explained

`r tab_ref("f_variance")` and `r fig_ref("f_variance")` put the features on one scale for five outcomes. One-way is the share of the variance explained by the feature alone; drop-one is the loss when the feature is left out of the model with all five. The two differ for the dependent features (Legal Representation, Vehicle type, Business use), whose one-way shares include part of their partner's effect.

Injury Severity leads for every outcome. For the claim size the order is Injury Severity (`r num(r2("log claim size", inj), 1)` points), Vehicle type (`r num(r2("log claim size", veh), 1)`), Age of Claimant (`r num(r2("log claim size", age), 1)`), Legal Representation (`r num(r2("log claim size", leg), 2)`) and Business use (`r num(r2("log claim size", bus), 2)`). For the settlement delay and the payment lag the order changes: Injury Severity, then Age of Claimant, then Legal Representation; Vehicle type and Business use explain nothing.

The features explain little of the variation between single claims: `r num(r2("log claim size", "additive", "r2"), 1)`% of the variance of the log claim size, `r num(r2("log settlement delay", "additive", "r2"), 1)`% for the log settlement delay and `r num(r2("log notification delay", "additive", "r2"), 2)`% for the log notification delay. The rest is the randomness of the generator. A small share of the variance does not mean a small effect on the portfolio: Injury Severity explains `r num(r2("log settlement delay", inj), 1)`% of the variance of the log settlement delay, yet the mean settlement delay of level 5 is `r num(lv(delay_level, inj, "5", "mean_setldel") / lv(delay_level, inj, "1", "mean_setldel"), 2)` times that of level 1. The saturated model (one mean per combination) explains hardly more than the additive one, so the features act almost multiplicatively, as designed. The last rows are the check: the features explain none of the variance of the claim size before the features.

```{r f-variance}
terms <- c(leg, inj, age, veh, bus)
variance_tab <- do.call(rbind, lapply(unique(variance$outcome), function(o) {
  v <- variance[variance$outcome == o, ]
  data.frame(outcome = o,
             measure = c("one-way", "drop-one"),
             rbind(100 * v$one_way[match(terms, v$term)],
                   100 * v$drop_one[match(terms, v$term)]),
             additive = c(NA, 100 * v$r2[v$term == "additive"]),
             saturated = c(NA, 100 * v$r2[v$term == "saturated"]))
}))
show(variance_tab, "f_variance",
     c("Outcome", "Measure", "Legal Representation", "Injury Severity",
       "Age of Claimant", "Vehicle type", "Business use",
       "All five (additive)", "All combinations (saturated)"),
     digits = c(0, 0, 3, 3, 3, 3, 3, 3, 3))
```

```{r fig-f-variance, out.width = "6.3in", fig.cap = fig_cap("f_variance")}
include_graphics(feat_fig("07 variance explained"))
```

# Impact of the features on the development and the reserves

## Payment pattern

`r tab_ref("f_pattern")` and `r fig_ref("f_pattern")` give the payment pattern of every level: the cumulative share of its nominal ultimate paid by development year, all accident years together. `r tab_ref("f_timing")` adds the amount-weighted mean payment lag, the mean time from the occurrence to the payment weighted by the nominal amounts.

Three features change the timing. Injury Severity: the lag rises from `r num(lv(timing, inj, "1", "lag"), 2)` years (level 1) to `r num(lv(timing, inj, "5", "lag"), 2)` years (level 5), and the share paid in the accident year falls from `r num(lv(timing, inj, "1", "dev_1"), 1)`% to `r num(lv(timing, inj, "5", "dev_1"), 1)`%. Age of Claimant: the band 0-15 has a lag of `r num(lv(timing, age, "0-15", "lag"), 2)` years against `r num(lv(timing, age, "over 65", "lag"), 2)` for over 65. Legal Representation: unrepresented claims are paid fastest (`r num(lv(timing, leg, "N", "lag"), 2)` years; `r num(lv(timing, leg, "N", "dev_2"), 1)`% by the end of development year 2), which is partly the effect of Injury Severity 1. Vehicle type and Business use leave the pattern where it is: the lags of Passenger and Light Commerical are `r num(lv(timing, veh, "Passenger", "lag"), 2)` and `r num(lv(timing, veh, "Light Commerical", "lag"), 2)` years. The patterns of Medium goods and Heavy goods are irregular because few large claims carry them.

```{r f-pattern}
show(pattern[, c("feature", "level",
                 paste0("dev_", c(1, 2, 3, 4, 5, 7, 10, 15, 20)))],
     "f_pattern",
     c("Feature", "Level", paste("DY", c(1, 2, 3, 4, 5, 7, 10, 15, 20))),
     digits = c(0, 0, rep(1, 9)))
```

```{r fig-f-pattern, out.width = "6.3in", fig.cap = fig_cap("f_pattern")}
include_graphics(feat_fig("09 payment pattern"))
```

```{r f-timing}
show(timing, "f_timing",
     c("Feature", "Level", "% of the ultimate", "Payment lag (years)",
       "Median lag of a claim", "Paid % by DY 1", "DY 2", "DY 3", "DY 5",
       "DY 10"),
     digits = c(0, 0, 2, 2, 2, 1, 1, 1, 1, 1))
```

## Exposure and outstanding

`r tab_ref("f_exposure")` and `r fig_ref("f_shares")` are the reserving view of the features: the weight of each level in the claims, in the cost and in the true outstanding. A level is over-represented in the outstanding when its claims are large or slow. Injury Severity 5 is `r num(lv(exposure, inj, "5", "claims_pct"), 2)`% of the claims and `r num(lv(exposure, inj, "5", "outstanding_pct"), 1)`% of the outstanding; Injury Severity 1 is `r num(lv(exposure, inj, "1", "claims_pct"), 1)`% of the claims and `r num(lv(exposure, inj, "1", "outstanding_pct"), 1)`% of the outstanding. The paid ratio falls from `r num(lv(exposure, inj, "1", "paid_pct"), 1)`% (level 1) to `r num(lv(exposure, inj, "5", "paid_pct"), 1)`% (level 5). Light Commerical vehicles and Business use Y are over-represented because of their size only: their paid ratios (`r num(lv(exposure, veh, "Light Commerical", "paid_pct"), 1)`% and `r num(lv(exposure, bus, "Y", "paid_pct"), 1)`%) are those of the portfolio (`r num(p[["paid to the valuation date (% of the ultimate)"]], 1)`%).

```{r f-exposure}
show(exposure, "f_exposure",
     c("Feature", "Level", "Claims", "% of claims", "% of cost",
       "% of ultimate", "% of outstanding", "Paid %", "Mean ultimate",
       "Outstanding / claims"),
     digits = c(0, 0, 0, 2, 2, 2, 2, 1, 0, 2))
```

```{r fig-f-shares, out.width = "4.72in", fig.cap = fig_cap("f_shares")}
include_graphics(feat_fig("03 shares by level"))
```

`r tab_ref("f_open")` splits the outstanding of each level into open and unreported claims. The levels differ far more in the outstanding per open claim (`r num(min(open_level$os_per_open_claim))` to `r num(max(open_level$os_per_open_claim))`) than in the share of their outstanding that is on unreported claims. `r tab_ref("f_os_injury")` shows the outstanding by accident year and Injury Severity: in the older accident years it is almost all on the high severities, while in accident year 20 every severity still has claims to pay.

```{r f-open}
show(open_level, "f_open",
     c("Feature", "Level", "Open claims", "% of the level's claims",
       "Not yet reported", "Outstanding, open",
       "Outstanding, not yet reported", "% of outstanding not yet reported",
       "Outstanding per open claim"),
     digits = c(0, 0, 0, 1, 0, 1, 1, 1, 0))
```

```{r f-os-injury}
show(os_injury, "f_os_injury",
     c("AY", paste("Injury", 1:6), "Total"),
     digits = 1)
```

## Chain ladder by level

`r tab_ref("f_cl_factors")` and `r fig_ref("f_factors")` give the chain-ladder factors of each level's own observed triangle. They are the quantity the NN chain ladder of Wuthrich (2018) models as a function of the features. The first factor rises with Injury Severity from `r num(lv(factors_level, inj, "1", "f_1"), 3)` to `r num(lv(factors_level, inj, "5", "f_1"), 3)` and is `r num(lv(factors_level, age, "0-15", "f_1"), 3)` for the age band 0-15 against `r num(lv(factors_level, age, "over 65", "f_1"), 3)` for over 65; for Passenger and Light Commerical it is the same (`r num(lv(factors_level, veh, "Passenger", "f_1"), 3)` and `r num(lv(factors_level, veh, "Light Commerical", "f_1"), 3)`). The last three columns show how far a level can be trusted: Heavy goods has as few as `r num(lv(factors_level, veh, "Heavy goods", "min_claims"))` claims in an accident year and `r num(lv(factors_level, veh, "Heavy goods", "zero_cells"))` of its 210 observed cells are empty, and the standard deviation of its individual first factors is `r num(lv(factors_level, veh, "Heavy goods", "sd_f_1"), 1)`.

```{r f-cl-factors}
show(factors_level, "f_cl_factors",
     c("Feature", "Level", paste0("f", 1:8), "Paid % in DY 1",
       "SD of individual f1", "Fewest claims in an AY", "Empty cells"),
     digits = c(0, 0, rep(3, 8), 1, 2, 0, 0))
```

```{r fig-f-factors, out.width = "6.3in", fig.cap = fig_cap("f_factors")}
include_graphics(feat_fig("10 chain-ladder factors"))
```

`r tab_ref("f_cl_reserves")` and `r fig_ref("f_cl_error")` apply Mack's chain ladder to each level's own triangle. The reserves of the large levels are within a few percent of the truth; those of the small or slow levels are not (Injury Severity 5 `r num(lv(cl_level, inj, "5", "error_pct"), 1)`%, Legal Representation N `r num(lv(cl_level, leg, "N", "error_pct"), 1)`%, Medium goods `r num(lv(cl_level, veh, "Medium goods", "error_pct"), 1)`%), and for Heavy goods the sqrt(msep) is larger than the reserve itself. These errors are single draws of one simulation, not biases.

```{r f-cl-reserves}
show(cl_level, "f_cl_reserves",
     c("Feature", "Level", "Paid", "Chain-ladder reserve", "sqrt(msep)",
       "True outstanding", "Error", "Error %", "Error / sqrt(msep)"),
     digits = c(0, 0, 1, 1, 1, 1, 1, 1, 2))
```

```{r fig-f-cl-error, out.width = "4.72in", fig.cap = fig_cap("f_cl_error")}
include_graphics(feat_fig("12 chain-ladder reserves by level"))
```

`r tab_ref("f_cl_split")` answers whether a split by one feature brings the total closer to the truth. It does not in this simulation: the totals range from `r num(min(cl_split$CL_reserves[-1]), 1)` to `r num(max(cl_split$CL_reserves[-1]), 1)` million against `r num(cl_split$CL_reserves[1], 1)` million for the whole triangle and the true `r num(cl_split$true_reserves[1], 1)` million. The differences between the rows are small against the sqrt(msep) of `r num(cl_split$msep_sqrt_whole_triangle[1], 1)` million, so the table cannot rank the splits; that would need repeated simulations. What the split does change is the sum of the absolute errors of the segments, which is the price of thinner triangles.

```{r f-cl-split}
show(cl_split[, c("split", "segments", "CL_reserves", "true_reserves",
                  "error", "error_pct", "sum_abs_errors")],
     "f_cl_split",
     c("Split", "Triangles", "Sum of the reserves", "True outstanding",
       "Error", "Error %", "Sum of absolute errors"),
     digits = c(0, 0, 1, 1, 1, 2, 1))
```

# What this means for the models

- The whole triangle is well behaved: no drift in the mix, a stable pattern, inflation constant in calendar time. The chain ladder is hard to beat on the total reserve of this portfolio, and one simulation cannot show a difference of a percent or two between two methods.
- The features carry real information about the development. Injury Severity, the age band 0-15 and Legal Representation change the payment pattern and the chain-ladder factors by more than their noise; Vehicle type and Business use change the size of the claims but not their timing. A model with features can therefore be expected to improve the reserves of groups of claims rather than the total.
- The feature space is thin. Splitting the triangle by one feature already gives noisy reserves for the small levels, and the cells of the NN chain ladder hold a median of `r num(cmb[["median number of claims per cell"]])` claims. The gain from features has to come from pooling across levels, which is what the neural networks do.
- `r num(100 * st_total$os_unreported / (st_total$os_open + st_total$os_unreported), 1)`% of the true outstanding is on claims whose features are not known at the valuation date.
- Effects of the features must be measured all else equal. The one-way tables are confounded for Legal Representation, Vehicle type and Business use, and effects on delays estimated on closed claims are biased towards no effect.

# Appendix: tables, figures and how to reproduce them

Run from the project root, in this order (about half a minute and one minute):

1. `analysis/00_claim-simulation/SynthETIC claims description.R`
2. `analysis/00_claim-simulation/SynthETIC feature impact.R`
3. knit this file (`thesis/claude_explainer-synthetic-claims-data-and-feature-impact.Rmd`).

The Word format is that of `project_folder_guide.docx`: `thesis/guide_reference.docx` holds its styles (built by `thesis/make_guide_reference.R`) and `thesis/guide_tables.lua` gives every table the guide's table style.

Tables 2 to 20 are the files `synthetic_data_t01_...csv` to `synthetic_data_t19_...csv` in `output/tables/00_claim-simulation/data-description/`, in the same order. Tables 21 to 47 are `synthetic_features_t01_...csv` to `synthetic_features_t27_...csv` in `output/tables/00_claim-simulation/feature-impact/`, in the same order except that the two tables on the variance explained (t18 and t19) appear as Tables 39 and 38. The tables in the files are not rounded. The figures are the PNG files in `output/figures/00_claim-simulation/data-description/` and `feature-impact/`; their file names start with "SynthETIC data Fig" and "SynthETIC features Fig".

The functions are in `R/claims_description.R` (portfolio at the valuation date, relativity of a combination, chain-ladder factors, Cramer's V, Gini coefficient, log-linear effects, variance explained) and `R/plots_claims_description.R` (figures), with checks in `tests/testthat/test-claims_description.R`.

References: Avanzi, Taylor, Wang and Wong (2021), SynthETIC: an individual insurance claim simulator with feature control, Insurance: Mathematics and Economics 100. Al-Mudafer, Avanzi, Taylor and Wong (2021), Stochastic loss reserving with mixture density neural networks. Gabrielli and Wuthrich (2018), An individual claims history simulation machine, Risks 6(2):29. Mack (1993), Distribution-free calculation of the standard error of chain ladder reserve estimates, ASTIN Bulletin 23. Wuthrich (2018), Neural networks applied to chain-ladder reserving, European Actuarial Journal 8:407-436. The publication details are quoted as in the project's scripts and were not re-checked for this report.
